---
title: "lecture3_extra"
author: "Kevin Allen"
date: "November 26, 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

## Introduction

Here are some extra exercises to gain more experience using R for data analysis.

We will use the same T-maze data set. In this data set, every row is trial. Each trial has 2 phases: the sample and choice phases.
To obtain a food reward on the choice phase, the mouse had to alternate between the two goal arms. For example, if the mouse entered the R goal arm during the sample phase, it then had to enter the L goal arm during the choice phase.

We will try to understand why some mice performed poorly on the rewarded alternation task.

## Load data from a file

```{r review_load}
myFile="~/repo/dataNeuroCourse/dataSets/tmaze.csv" 
df<-read_csv(myFile)
df<-mutate(df, correct = sample != choice)
```

## Exploratory data analysis

One hypothesis is that some mice had a bad performance because they always visited the same goal arm instead of alternating. 

How could we test this?

We would first need a score per mouse telling us if the mouse always choose the same arm?


## Break down your programming task into smaller tasks.

Strategy 1

* Which column contains relevant information?
* How can we go from the values in this column to a statistic (number)?
* Once you have the information, choose a plot?

Strategy 2

* Imagine which plot would convey your message best (draw some on paper)
* Find out which steps are needed to produce this graph.

***

In our case, the score would be calculated from the column "choice".

To create a statistic, we need to transform a `chr` to something we can count (e.g., `dbl` or `lgl`).
Change `R` and `L` to `0` and `1` using `ifelse`.

```{r noAlternation1}
df <- df %>% mutate(turn = ifelse( choice == "R" , 0, 1))
head(df,n=5)
```

***

The "turn" column now tells us whether the animal turned right(0) or left(1).

Now summarize this information as a new variable for each mouse on each block.

Also get the performance from the `correct` column, as done last week.

***

```{r noAlternation2, out.width = "800px"}
df1 <- df %>% 
  group_by(mouse,block) %>% 
  summarise(nonAlternance = mean(turn),performance = mean(correct))
df1
```

***

Our nonAlternance score is 0 or 1 if the mouse always goes to the right or left, respectively.

To know if mice prefer left or right turns, plot the distribution of nonAlternation.

*** 

```{r noAlternation3,out.width = "500px"}
df1 %>% 
  ggplot() +
  geom_histogram(mapping = aes(x = nonAlternance),binwidth = 0.1)
```

The distribution is not centered on 0.5. This means that our mice had a small tendency to turn left in our experimental room.

***

Plot the performance and nonAlternation to see how the two variables are related.

***

```{r noAlternation4,out.width = "500px"}
df1 %>% 
  ggplot() +
  geom_point(mapping = aes(x = nonAlternance, y = performance))
```

Many points are on top of each other. This is called [overplotting](https://www.data-to-viz.com/caveat/overplotting.html). 

Try solving this by setting `position="jitter"` or `alpha=0.1`.

*** 

```{r noAlternation5,out.width = "400px"}
df1 %>% 
  ggplot() +
  geom_point(mapping = aes(x = nonAlternance, y = performance),
             position = "jitter") +
  geom_smooth(mapping = aes(x = nonAlternance, y = performance),
              method = "loess",formula = 'y ~ x')
```

The geom_smooth helps you see the pattern in the data.

If a mouse always goes to the same goal arm (e.g., always left), it is not alternating. Therefore, the performance goes down.


***

Plot the nonAlternance on every block for every mice, with our chance level line at 0.5.

*** 

```{r nonAlternance6,out.width = "400px"}
df1 %>% 
  ggplot() +
  geom_line(mapping=aes(x=block,y=nonAlternance)) +
  geom_hline(yintercept = 0.5, alpha=0.5) +
  facet_wrap(~mouse)
```

Most mice had a moderate tendency to turn left (1), but one (Mn7735) had a persistent tendency to turn right (0).

Can you add the performance to this graph?

***


```{r nonAlternance8,out.width = "400px"}
df1 %>% 
  ggplot() +
  geom_line(mapping=aes(x=block,y=performance) ) +
  geom_line(mapping=aes(x=block,y=nonAlternance),color="blue",alpha=0.3) +
  geom_hline(yintercept = 0.5, alpha=0.5) +
  facet_wrap(~mouse)
```

***

Here we add label to each point.

```{r nonAlternance7,out.width = "400px"}
df %>% group_by(mouse) %>% 
  summarise(nonAlternance = mean(turn),performance = mean(correct)) %>% 
  ggplot(mapping=aes(x=nonAlternance,y=performance)) +
  geom_point() +
  geom_text(aes(label=mouse),hjust=-0.1,vjust=0.5) +
  xlim(0.2,0.8)
```

So the poor performance of Mn7735 is associated with a tendency to turn right during the choice phase of the trials.
