---
title: "Data science and analysis in Neuroscience"
author: "Kevin Allen"
date: "December 3, 2019"
output:
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

## Today's plan

1. Combining information from several data frames
2. Common statistical procedures with R
  + Correlation 
  + t-test
  + ANOVA
  + Common assumptions and common errors


## Combining information from several data frames

Load the tmaze data set for a few more exercises.

```{r load}
myFile="~/repo/dataNeuroCourse/dataSets/tmaze.csv" 
df<-read_csv(myFile)
df<-mutate(df, correct = sample != choice)
```


## Combining information from several data frames

In most project, you need to work with several tables. 

Relations are defined between a pair of tables.

dplyr has several functions to do this.


## Combining information from several data frames

Let's assign a genotype to each mouse in a new data frame.

```{r relational1}
dfGeno<-tibble(mouse=c("Mn4656","Mn848","Mn4672","Mn4673",
                           "Mn7712","Mn7735","Mn829"),
           genotype=c("wt","wt","wt","wt",
                      "ko","ko","ko"))
dfGeno
```

***

How is `df` related to `dfGeno`?

```{r relational2}
colnames(df)
colnames(dfGeno)
colnames(dfGeno)[colnames(dfGeno) %in% colnames(df)]
```
`mouse` is a **key**, a variable that connect a pair of tables.

***

* A **primary key** uniquely identifies an observation in its table.

```{r relational3}
dfGeno %>% 
  count(mouse)

```

***

* A **foreign key** uniquely identifies an observation in *another* table.

```{r relational4}
df %>% 
  count(mouse)
```

## Mutating joins

* It first matches observation by their keys.
* Then copies across variables from one table to the other.

```{r relational5}
df_join <- df %>% 
  left_join(dfGeno,by="mouse") # match with mouse
df_join
```
We now have an additional variable (genotype) in df.

For more information: [Relational data and dplyr](https://r4ds.had.co.nz/relational-data.html)


Time to use `git` to `commit` and `push` your changes to your online repository.


## Common statistics with R

1. Correlation 
2. t-test
3. Wilcoxon tests
3. ANOVAs
4. Common assumptions and common errors

## Warnings

Always make sure that you understand the assumptions associated with each test.

A p value has no real meaning if the assumptions of the test are violated.

## Reading for this week

[Ten common statistical mistakes (eLife, 2019)](https://elifesciences.org/articles/48175?gclid=Cj0KCQiAwf39BRCCARIsALXWETyV3Dhu7jGevlenU1oh2zzLo812jOvAeparT0UYPEY9hPOFI34ECUQaAkBEEALw_wcB)
